---
title: "Danish diabetes data"
author: "Henrik Vitus Bering Laursen"
date: "2025-06-01" 
categories: 
  - diabetes
  - joins
  - survival analysis
draft: true
---

## TODO

- cox ass test
- log-rank model test for dif
- 

## Purpose

I remember using the `Epi::steno2` data-sets when taking a short course on multistate survival analysis with the creator of the package, [Bendix Carstensen](https://bendixcarstensen.com/).

That was quite fun but also daunting to be sitting in a room full of mathematicians and people who actually knew the nitty gritty of survival analysis and the math behind it.

I think I would just want to do something very basic here, though:

- A) Playing around with joining the different data-sets
- B) Examine the data-sets and visualize the contents
- C) Creating a simple Cox Proportional Hazards model, comparing the risk of cardiovascular event between the two treatment groups in the data-set
  - Probably also controlling for some important factors
- D) Deliver some sort of conclusion about these data
  
So lets start looking at it!

## Load

The data-sets can be loaded directly from the `Epi` library, through `data()`

```{r}
#| label: Loading Libraries
#| echo: false
library(Epi)
library(tidyverse)
library(survival)
library(survminer)
```

```{r}
#| label: loading steno2 data
data("steno2")
data("st2clin")
data("st2alb")
```

Alrighty now that they are loaded lets look at their structure. **Ofcourse, I am completely ignoring the description in the documentation in the `Epi()` package, which describes how the data-sets are structured.**

```{r}
#| label: Structure of data-sets
str(steno2)
str(st2clin)
str(st2alb)
```

It looks like `steno2` and `st2clin` are one row per subject, and `st2alb` is multiple, but that doesnt make sense given the amount of rows. We can see how many id's, the unqiue identifier for each person, are present, and if there is one row per subject it should be close to the amount of subjects.

```{r}
#| label: unique IDs

unique_new <- function(df) {df |> group_by(id) |> summarise(n = sum(id)) |> slice_max(n)}

map(list(steno2,st2clin,st2alb),unique_new)

```

Alright. So it is clear that only `steno2` has one row per subject because the number of id's that occur equal the number of unique id's.

It just looks like `st2clin` is one row per subject because of how it is arranged. 

```{r}
#| label: st2clin records

st2clin |> arrange(chol) |> head(4)
```

### A) Joining

Now, first, what would the purpose be of joining these data-sets together?

First of all, the `st2clin` and `st2alb` data-sets do not have the categories for each id from the `steno2` data-set.

## Output
### B) Table

Ofcourse, the good old `summary()` function might be worth to do to get a good overview of the three data-sets. First we do `steno2` demographics.

```{r}
#| label: steno2 demographics

summary(steno2)

```
But I kinda want to mainly do graphs here.

### B) Graphs

#### Some distributions

Now, I love me some functions, and I feel like seeing the distribution of all the different date variables via histograms.

```{r}
#| label: first attempt, going over date vars
plot_histo_spam <- function(df,histovar) {
out <- ggplot(df) + 
  geom_histogram(mapping = aes( {{histovar}} ))

  return(out)
}

plot_histo_spam(steno2,doBth)
plot_histo_spam(steno2,doDM)
plot_histo_spam(steno2,doBase)

```

But this should be doable in a smarter way. A `facet_wrap()` over each of the date variables..

```{r}
#| label: second attempt, going over date vars

plot_histospam_faceted <- function(data, prefix = "do") {
  data |> 
    select(starts_with(prefix)) |>             
    pivot_longer(everything()) |> # everything() turns varnames into a column, and their values into a separate column
    ggplot(aes(value)) +
      geom_histogram(colour = "white") +
      facet_wrap(~ name, scales = "free_x") +   # one panel per variable
      labs(x = NULL) 
}

plot_histospam_faceted(steno2)
plot_histospam_faceted(st2clin)
plot_histospam_faceted(st2alb)

```

Now those are some fine basic plots! 

For `steno2`, it shows some slight right skewed `doBase`, and left-skewed `doDM` and `doEnd`. for the two `st2*` data-sets, it shows quite a right skew.

**Basically**:

- most of the signups to the trial occurred in the beginning
- most of the CVD (cardiovascular) and death events occurred evenly-ish
- most of the diabetes diagnoses occurred late
- not much ESRD (end-stage renal disease) occurred

#### Distributions by sex

Now I want to figure out how to do it while grouping the histograms by sex.

```{r}
#| label: histograms, faceted, sex

plot_histospam_faceted_sex <- function(data, prefix = "do") {
  data |> 
    select(sex, starts_with(prefix)) |>             
    pivot_longer(-sex, 
                 names_to = "name",
                 values_to = "value") |> 
    ggplot(aes(value, fill = sex)) +
      geom_histogram(
        position = "dodge",
        colour = "white") +
      facet_wrap(~ name, scales = "free_x") +   # one panel per variable
      labs(x = NULL) 
}

plot_histospam_faceted_sex(steno2)
```

Ok. Not much of a difference at a glance. 

#### Distributions by treatment groups

Perhaps it would be more relevant to see if there was some kind of difference between the two intervention groups?

```{r}
#| label: histograms, faceted, allocation group

plot_histospam_faceted_allo <- function(data, prefix = "do") {
  data |> 
    select(allo, starts_with(prefix)) |>             
    pivot_longer(-allo, 
                 names_to = "name",
                 values_to = "value") |> 
    ggplot(aes(value, fill = allo)) +
      geom_histogram(
        position = "identity",
        alpha    = 0.5,
        colour = "white") +
      facet_wrap(~ name, scales = "free_x") +   # one panel per variable
      labs(x = NULL) 
}

plot_histospam_faceted_allo(steno2)
```

Ok so already here its possible to see some apparent differences between the two groups.

- the `Conv` allocation group, which must be those allocated to conventional treatment, whatever that was, have more events of CVD, especially of the third CVD event. As well as earlier diabetes mellitus (DM) diagnoses. This is probably an indication of the future results of the survival analysis.
- `Int` group, likely the intervention group, seems to have less deaths and later DM diagnoses

**Remember, these are two groups of equal number**, which can make it easier to conclude that there is a difference, at least initially and visually.

Now. With the help of some AI, I made the function below which can produce different types of `facet_*` plots , but I choose the `facet_grid`, because its interesting to look at.

```{r}
#| label: histograms, faceted, complicated

plot_histospam_faceted_allo_comp <- function(
    data,
    prefix      = "do",
    bins        = 30,
    overlay     = TRUE,     
    facet_rows  = FALSE      
) {
  data |> 
    select(allo, starts_with(prefix)) |>         
    pivot_longer(-allo, names_to = "variable",
                       values_to = "value") |>   
    ggplot(aes(value, fill = allo)) +
      geom_histogram(
        bins      = bins,
        position  = if (overlay) "identity" else "dodge",
        alpha     = if (overlay) 0.5 else 1,      # translucent if overlaid
        colour    = "white"
      ) +
      { if (facet_rows)
          facet_grid(rows = vars(allo), cols = vars(variable),
                     scales = "free_x")
        else
          facet_wrap(~ variable, scales = "free_x")
      } +
      labs(x = NULL, fill = "Sex") +
      theme_minimal()
}

plot_histospam_faceted_allo_comp(steno2, facet_rows = TRUE)

```


### C) Model

To perform a basic Cox Proportional Hazards model, we can use the `survival` package.

I want to do the following, just because that's what I learned in the olden PhD studying days, mainly from the Basic Biostats courses from Aarhus University:

- Survival set the data
- Visualise follow-up
- Plot Kaplan Meier survival function
- Cox model
  - Estimate
  - Observed vs fitted survival curves
  - Testing proportional hazards assumption
  
Let's just quickly and without much thinking (thats always good), decide what the start, stop, outcome, and other variables should be. 

- data: `steno2`
  - comparison: death `doDth` for intervention groups `allo`
  - start date: `doBase`
  - censor date: `doEnd`
  - outcome event: `doDth`
  - exit of study: earliest of `doEnd` and `doDth`

So, super basic, and does not take anything into account, such as competing events (`doCVD*` or `doESRD`) or baseline differences in risk of CVD (`baseCVD`). 



#### Survival set the data

First we create the needed variables and check it.

```{r}
#| label: survival setting

steno2 <- steno2 |> 
  mutate(
    doExit = coalesce(doDth, doEnd),
    time_days = difftime(
      coalesce(steno2$doDth, steno2$doEnd), 
      steno2$doBase, 
      units='days'),
    time_yrs = as.numeric(time_days / 365.25),
    event_num = if_else(!is.na(doDth),1,0),
    event = factor(event_num, levels = c(0, 1), labels = c("Censored", "Dead"))
  )

glimpse(steno2)
```

#### Visualise follow-up

Then, because I like it as an overview, a swimmer plot of the follow-up times.

simple first:

```{r}
#| label: swimplot followup simple
sp1 <- steno2 |> 
  mutate(id = fct_reorder(as.factor(id), time_yrs)) |>
  ggplot(aes(x = doBase, 
             xend = doExit,
             y = id)) +
  geom_segment(color = "grey") +
  geom_point(aes(x = doExit, shape = event)) +
  geom_point(aes(x = doDth, shape = event)) 
sp1
```

Looks good, but a tiny bit messy with the - although accurate - different starting times.

Lets do one with baseline time set to 0. This makes it easier to compare the follow-up times, which is more interesting for the analysis than calendar times.

The latter would probably be of more interest to an investigator trying to figure out when people signed up, or something.

```{r}
#| label: swimplot followup simple base is 0

sp2 <- steno2 |> 
  mutate(
    id = fct_reorder(as.factor(id), time_yrs),
    t0 = 0) |>
  ggplot(aes(x = t0, 
             xend = time_yrs,
             y = id)) +
  geom_segment(color = "grey") +
  geom_point(aes(x = time_yrs, shape = event, colour = event)) +
  theme_minimal()
sp2
```

Here we have the x-axis in years since baseline.

Now, I have a sense that `doExit` and `doDth` are the same, meaning that plots that use `doExit` end up having two shapes for subjects that died.

How can I make it so that there are different shapes for each event? I can just take the previous plot and add to it.


```{r}
#| label: swimplot different shapes for event
sp3 <- sp2 +
  scale_shape_manual(values = c(Censored = 17, Dead = 4)) +
  scale_colour_manual(values = c(Censored = "black", Dead = "red"))  
sp3
```
Ok that looks fine. Now I want to make the final plot for this and move on.

```{r}
#| label: swimplot followup final

sp4 <- sp3 +
  labs(
    x = NULL, 
    y = "Subject sorted by follow-up", 
    title = "Follow-up per subject, years since baseline") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    plot.background = element_blank(),
    panel.background = element_blank()
    )
sp4
```

A nice little plot showing an overview of the length of follow-up for the subjects in the study.

Not exactly sure how it is useful except for my own sake - I just like to know *visually* what my data looks like.

#### Disclaimer

I am not a statistician, so whatever is correct of below, is due to the wise words and lectures I have remembered from people who actually know what they are talking about. Many statistical results, especially when you wish to talk inference (X causes Y), require careful consideration of subjective priors, demographics, causal effects, and biases such as effect modifiers, confounders, selection bias, etc.

Now - with that out of the way: LETS CLAIM SOME THINGS!

#### Plot Kaplan Meier survival function

Now, less fiddling around with plots, and more looking at survival analysis.

The Kaplan meier estimator on a plot is wonderful for portraying differences in survival times. First step we have is to fit the data.

```{r}
#| label: surv fit

# Null, by treatment allocation, subset into sex
fit_null <- survfit(Surv(time_yrs, event_num) ~ 1, data = steno2)
fit_allo <- survfit(Surv(time_yrs, event_num) ~ allo, data = steno2)
fit_sex <- survfit(Surv(time_yrs, event_num) ~ sex, data = steno2)
fit_allo_sex <- survfit(Surv(time_yrs, event_num) ~ allo + sex, data = steno2)

# steno2_m <- steno2 |> filter(sex == "M") 
# steno2_f <- steno2 |> filter(sex == "F") 
# fit_allo_m <- survfit(Surv(time_yrs, event_num) ~ allo, data = steno2_m)
# fit_allo_f <- survfit(Surv(time_yrs, event_num) ~ allo, data = steno2_f)

# Summarising, showing just the first 10 years
summary(fit_null, times=c(0:10)) 
summary(fit_allo, times=c(0:10)) 
summary(fit_sex, times=c(0:10)) 
summary(fit_allo_sex, times=c(0:10)) 
```
Here we can see that after 10 years, there is already a lower survival probability for the conventional treatment group. And for females, it seems. But that highly depends on the intervention group.

But it is much better to visualize it with a kaplan meier plot, to really see the differences, and when - in terms of time since baseline - does the survival differ.

Shown below are plots for each of the different fits mentioned above, showing the null model, differences in survival between intervention groups for both sexes, and ditto, but subset for each sex.

```{r}
#| label: survplot KM

# Function to save some space
survplot_function <- function(fit,data){
  out <- ggsurvplot(
    fit, data = data,
    conf.int = TRUE,
    risk.table = TRUE,
    palette = "dark2",
    xlab = NULL, ylab = "Survival Probability",
    title = "Kaplan Meier Estimator for survival probability, in years"
    )
  
  return(out)
}

# Null
survplot_function(fit_null,steno2)

# allo
survplot_function(fit_allo,steno2)

# sex
survplot_function(fit_sex,steno2)

# allo sex
survplot_function(fit_allo_sex,steno2)

```

Thats a bit rough to look at, but provides plenty of information:

1. intervention group has higher survival probability
2. female sex seems to indicate having the highest survival probability, at least in intervention group
3. frequently overlapping confidence intervals, low probability of actual difference atleast in the first 10 years where most of the overlap occurs

For the second point, the lines cross twice, either because of the lower number of females or because of some unknown factor. When the lines cross its difficult to say whether there is any consistent survival difference with certainty.

But let's try to make it pretty - just with the help of the `survminer` package help - accessed in Rstudio via `?ggsurvplot`.

```{r}
#| label: survplot KM pretty, fewer

# allo
surv_p_allo_pretty <- ggsurvplot(
  fit = fit_allo, data = steno2,
  surv.median.line = "hv", # Add medians survival
  legend.title = "Allocation group",
  legend.labs = c("Intervention", "Conventional"),
  pval = TRUE,
  conf.int = TRUE,

  risk.table = TRUE,
  tables.height = 0.2,
  tables.theme = theme(axis.title.y = element_blank()), 
  
  palette = "dark2",
  ggtheme = theme_bw(),
  xlab = NULL, ylab = "Survival Probability",
  title = "Kaplan Meier Estimator for survival probability, in years"
  )

surv_p_allo_sex_pretty <- ggsurvplot(
  fit = fit_allo_sex, data = steno2,
  surv.median.line = "hv", # Add medians survival
  legend.title = "Allocation group, Sex",
  legend.labs = c("Int, F", "Int, M", "Conv, F", "Conv, M"),
  conf.int = TRUE,

  risk.table = TRUE,
  tables.height = 0.2,
  tables.theme = theme(axis.title.y = element_blank()), 
  
  palette = "dark2",
  ggtheme = theme_bw(),
  xlab = NULL, ylab = "Survival Probability",
  title = "Kaplan Meier Estimator for survival probability, in years"
  )

surv_p_allo_pretty
surv_p_allo_sex_pretty
```
Now slightly cleaned up, and with some median survival lines - the follow-up time at which 50 has experienced the event - added.

##### Test of differences

One way of testing the differences between two survival curves is the log-rank test. Although, as far as I remember, its not given much weight - especially compared to visual inspection coupled with careful thinking and analysis of the different factors that play into a survival analysis.

```{r}
#| label: Log-rank test
survdiff(Surv(time_yrs, event_num) ~ allo, data = steno2)
survdiff(Surv(time_yrs, event_num) ~ sex, data = steno2)
```
Seems there is a statistically significant difference between allocation groups and not sex. Kinda what I expected from the reading of the plots.

I have forgotten how to interpret the log-rank model for a fit with two different binary variables, so I will refrain from attempting that at the moment.

```{r}
#| label: CHUNK FOR INSPIRATION FROM PHD
# Log-rank test 
survdiff(Surv(T_E_Y, EVENT) ~ GROUP, data = df_tte_s_e)
survdiff(Surv(T_E_Y, EVENT) ~ GROUP, data = df_tte_s_l)

# loglog for proportional hazards
ggsurvplot(groupsurv_e, data = df_tte_s_e, fun = "cloglog")
ggsurvplot(groupsurv_l, data = df_tte_s_l, fun = "cloglog")

## Conclusion: Early, the assumption seems to be violated because GLP1 and DPP4 cross.
## Looks fine for late, except the very end (beginning?) where GLP1 and SGLT2 cross.

# Linearity (GROUP*LATE)?
  # for og unstratified
plot(predict(cox_a), residuals(cox_a, type = "martingale"), las = 1)
lines(smooth.spline(predict(cox_a), residuals(cox_a, type = "martingale")), col = "red")
  # early
plot(predict(cox_a_e), residuals(cox_a_e, type = "martingale"), las = 1)
lines(smooth.spline(predict(cox_a_e), residuals(cox_a_e, type = "martingale")), col = "red")
  # Late
plot(predict(cox_a_l), residuals(cox_a_l, type = "martingale"), las = 1)
lines(smooth.spline(predict(cox_a_l), residuals(cox_a_l, type = "martingale")), col = "red")

# Schoenfeld
cox.zph(cox_a)

par(mfrow=c(4,2))
plot(cox.zph(cox_a))
```


#### Cox model

Now we enter the part of survival analysis where we want to 

##### Estimate
##### Observed vs fitted survival curves
##### Testing proportional hazards assumption

## Summary / cleaned up / refactored

